Synthesizing Verified Components for Cyber Assured Systems Engineering \thanks{DISTRIBUTION STATEMENT A. Approved for public release.}

### 2022-03-29 paper/synthesis.tex
checkpointing; Section 7 proof story fleshed out

Synthesis maps from model and specifications to code. The SPLAT tool
traverses the system architecture looking for occurrences of high
assurance components specified by code contracts; for each such
occurrence it generates a CakeML program. More generally, SPLAT supports
three kinds of behavioral specification from which to generate
code. In increasing order of expressiveness, these are:

\paragraph{Regular expressions.} Often the well-formedness of data on connections
can be enforced by matching against a regular expression. In other
words, the data can be characterized by a \emph{regular language}. As
is well known, regular expressions can be translated to efficient
deterministic finite state machine (DFA)
implementations. In \cite{formal-filter-synth-langsec,case-verified-filter}
we report on the creation and application of filters from a verified
regular expression-to-DFA compiler.

\paragraph{Contiguity types\cite{contiguity-types}.}
A contiguity type, like regular expressions, is effective at
describing data at the network message level, \eg, as flat
strings. The notation is however more expressive, being able to
declare many kinds of so-called \emph{self-describing} messages, \ie,
those where message structure is dependent on data held elsewhere in
the message. Figure \ref{fig:filter-spec} gives a contiguity type
specification for the welll-formed waypoints of our example system. We
have formalized and verified a contiguity type matcher. The matcher
code has been used in our case studies.

\paragraph{AGREE code contracts.} When full computational power is needed, we can use the
expressive power of AGREE code contracts to write arbitrarily complex
filter or monitor operations, as we have already seen. Our formal
model of AGREE syntax and semantics is applied to extract code
contracts and map them to equivalent logic functions.

\noindent For each of these three approaches, we take the following steps to
arrive at an executable.

@@

Partiality is an important consideration: steps 2 and 3 above can
fail; the data might not be parseable or the wellformedness
computation could be badly written and fail at runtime. In such cases,
the filter should recover and yield control without passing the input
onwards. In these cases, the filter is behaving as it should, but we
must also guard against situations in which a \emph{correctly
specified} filter fails at runtime. This kind of defect arises when
the filter \emph{ought} to accept a message, but lack of resources
results in the filter failing to do so. For example, the parse of a
message might need more space than has been allocated; another example
could be if the time slice provided by the scheduler is too short for
the wellformedness computation to finish. Thus resource bounds need to
be included in the correctness argument. Some preliminary work on this
has been done in CakeML \cite{cakeml-space-cost}.

@@

Monitors are intended to track and analyze the externally visible
behavior of system components through time. Therefore, they tend to
require more extensive computational ability than filters. In
particular, our basic notion of a monitor is that it embodies a
relation over its input and output streams, and is able to access the
value of a stream at any earlier point in time, if necessary. Monitors
commonly use state to keep track of earlier values, unlike filters
which, for us, are typically stateless components. (However, there is
nothing in our approach that forbids stateful filters: they can be
realized by monitors.) A monitor specification is mapped by code
generation to a state transformation function of the following
abstract type:
\[
\konst{stepFn} : \mathit{input} \times \mathit{stateVars} \to \mathit{stateVars} \times \mathit{output}
\]

\subsubsection{Initialization}

A monitor may need to accumulate a certain minimum number of
observations before being able to make a meaningful assessment of
behavior. Until that threshold is attained, the monitor is essentially
in its \emph{initialization} phase. In order for correct code to
be generated, monitor specifications need to spell out the values of
output ports when in their initialization phases. For example, suppose a
monitor does some kind of differential assessment of inputs at
adjacent time slices, alerting when (say) the measured location of a
UAV at times $t$ and $t+1$ is such that the distance between the two
locations is unusually large. Such a monitor needs two measurements
before making its first judgement, but at the time of its first
output, only one measurement will have been made. The specification
must then explicitly state the correct value for the first output.

\subsubsection{Step function}

The \konst{stepFn} works as follows:

@@

@@

>>>>

Synthesis maps from model and specifications to code. The SPLAT tool
traverses the system architecture looking for occurrences of high
assurance components specified by code contracts; for each such
occurrence it generates a CakeML program. SPLAT supports three kinds
of behavioral specification from which to generate code. In increasing
order of expressiveness, these are: regular expressions, contiguity
types, and AGREE specifications. In earlier work we have described how
regular expressions and contiguity types are used as specification
languages for verified filter implementations. In the present paper,
we extend this \emph{specification oriented} approach to the
Turing-complete specifications of AGREE.  We give a short review of
that previous work before going into detail on AGREE.

\paragraph{Regular expressions.} Often the well-formedness of data on connections
can be enforced by matching against a regular expression. In other
words, the data can be characterized by a \emph{regular language}. As
is well known, regular expressions can be translated to efficient
deterministic finite state machine (DFA)
implementations. In \cite{formal-filter-synth-langsec,case-verified-filter}
we report on the creation and application of table-based DFA filters
from a verified regular expression-to-DFA compiler. The latter paper
offers a fully-worked out example synthesizing a regular expression
filter and formally proving that it meets an infinitary behavioral
specification.

\paragraph{Contiguity types\cite{contiguity-types}.}
Contiguity types, similar to regular expressions, are effective at
describing data at the network message level, \eg, as flat
strings. The notation is however much more expressive, being able to
declare many kinds of so-called \emph{self-describing} messages, \ie,
those where message structure is dependent on data held elsewhere in
the message. Figure \ref{fig:filter-spec} gives a contiguity type
specification for the welll-formed waypoints of our example system. We
have formalized and verified a contiguity type matcher, which has been
extensively applied throughout our case studies.

\subsection{Compiling from code contracts}

When full computational power is needed, we can use the expressive
power of AGREE code contracts to write arbitrarily complex filter or
monitor operations, as we have already seen. Our formal model of AGREE
syntax and semantics is applied to extract code contracts and map them
to equivalent logic functions. In particular, given a code contract with
functionality ${\Eqs}\cdot G$, input ports $\mathbb{I}$, state
variables $\mathbb{S}$, and output ports $\mathbb{O}$, the semantics
given in Section \ref{sec:code-contracts} can be applied to create a
function
\[
 \konst{strmStep} : \mathbb{N}_0 \to \mathit{Env} \to \mathit{Env} = \sem{{\Eqs}\cdot G}
\]
Assume ${\Eqs}\cdot G$ is \emph{Pascalish}, \ie, sequentially
imperative. Then we can direcctly create a corresponding logic function
\[
\konst{stateStep} : \mathit{input} \times \mathit{state} \to \mathit{state} \times \mathit{output}
\]
that takes input and state as parameters and returns an updated state
and output. Thus the environment $\mathit{Env}$ holding all previous
values of each variable can be replaced by \emph{parameters}, each of
which is able to hold only one data element. Assuming
$\mathit{Env}' = \konst{strmStep}\;t\; \mathit{Env}$, we can formally show that
\[
\vdash (\mathit{Env}' \vert_{\mathbb{S}},
 \mathit{Env}' \vert_{\mathbb{O}}) =
\konst{stateStep} (\mathit{Env}\vert_{\mathbb{I}(t)},
                   \mathit{Env}\vert_{\mathbb{S}(t)})
\]
In other words, projecting out the state and output components from
$\konst{strmStep}\;t\; \mathit{Env}$ equals the result of
applying \konst{stateStep} on the input and state components of
$\mathit{Env}$ at time $t$. This theorem connects the stream semantics
of the code contract to a pure logic function over the values
appearing in the stream. From this point we can use tools in the
CakeML eco-system to jump from the setting of higher order logic to
CakeML abstract syntax \cite{cakeml-translator}. Thus we have moved
from stream semantics to logic function to CakeML evaluation semantics
in a property-preserving way. Once in CakeML, further necessary
features can be added, namely imperative variables for storing state
between component invocations, also invocations of the (verified) foreign
function interface of CakeML. Finally, the CakeML compiler can be
applied to the program. It delivers an executable which can be shown
to preserve the original contract.

%% where the $\mathit{state}$ parameters to the function are just the
%% variables defined in ${\Eqs}$, obtained by projecting out the

%%  = \mathit{Env}\vert_{\mathbb{I}}$ $\mathit{input}
%%  = \mathit{Env}\vert_{\mathbb{I}}$ $\mathit{input}
%%  = \mathit{Env}\vert_{\mathbb{I}}$

\subsection{Generating CakeML}
For any of the three specification languages, we take the following general steps to
arrive at an executable.

@@

Partiality is an important consideration: steps 2 and 3 above can
fail; the data might not be parseable or the wellformedness
computation could be badly written and fail at runtime. In such cases,
the filter should recover and yield control without passing the input
onwards. In these cases, the filter is behaving as it should, but we
must also guard against situations in which a \emph{correctly
specified} filter fails at runtime. This kind of defect arises when
the filter \emph{ought} to accept a message, but lack of resources
results in the filter failing to do so. For example, the parse of a
message might need more space than has been allocated; another example
could be if the time slice provided by the scheduler is too short for
the wellformedness computation to finish. Thus resource bounds need to
be included in the correctness argument. Some preliminary work that
could support formal proofs of such bounds has been done for
CakeML \cite{cakeml-space-cost}.

@@

Monitors are intended to track and analyze the externally visible
behavior of system components through time. Therefore, they tend to
require more extensive computational ability than filters. In
particular, our basic notion of a monitor is that it embodies a
relation over its input and output streams, and is able to access the
value of a stream at any earlier point in time, if necessary. Monitors
commonly use state to keep track of earlier values, unlike filters
which, for us, are typically stateless components. (However, there is
nothing in our approach that forbids stateful filters: they can be
realized by monitors.) A monitor specification is mapped by code
generation to a state transformation function of the following
abstract type:
%
\[
\konst{stepFn} : \mathit{input} \times \mathit{stateVars} \to \mathit{stateVars} \times \mathit{output}
\]
The \konst{stepFn} works as follows:
\begin{enumerate}

@@

\subsection{Incorporation into system execution}

@@

%% \subsubsection{Initialization}

%% A monitor may need to accumulate a certain minimum number of
%% observations before being able to make a meaningful assessment of
%% behavior. Until that threshold is attained, the monitor is essentially
%% in its \emph{initialization} phase. In order for correct code to
%% be generated, monitor specifications need to spell out the values of
%% output ports when in their initialization phases. For example, suppose a
%% monitor does some kind of differential assessment of inputs at
%% adjacent time slices, alerting when (say) the measured location of a
%% UAV at times $t$ and $t+1$ is such that the distance between the two
%% locations is unusually large. Such a monitor needs two measurements
%% before making its first judgement, but at the time of its first
%% output, only one measurement will have been made. The specification
%% must then explicitly state the correct value for the first output.

% \subsubsection{Step function}
